Road Segment Prioritization for Bicycle Infrastructure

### 2021-01-19 README.Rmd
Updates from meeting

<!-- - Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization -->
- Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them
<!-- - Should we use other cities for the sake of comparison (compare communities, road types etc) -->
- How to add tables properly (they are now added using latex code which only shows up when outputting to pdf)
- Footnotes are sometimes split between pages. How do I keep them on the same page?
<!-- - What is the best way to write pseudocode (Steps for Algorithm 1 and 2) -->

@@

*Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. They account for the effect of cycling lanes on car traffic, and attempt to maximize utilization of said lanes with minimal impact on car travel times. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links^[*link* refers to a road segment throughout this research] is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all origin-destination (OD)^[*Origin-Destination (OD)* data is normally obtained from a census. In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between Middle layer Super Output Area (MSOA) zones, by mode of travel. MSOAs have an average population of just over 8000 [@ofn2018population].] pairs need to be connected by roads that meet the desired LOS. A limitation of these problem formulations do not explicitly solve for continuity, which is dealt with using either a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], a constraint on maximum deviation from shortest paths [@duthie2014optimization], or a discontinuity penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

@@

\noindent where **d** and **s** are the distance and slope respectively for the OD pair. The authors use square and square-root distance terms "to capture the non-linear impact of distance on the likelihood of cycling", and interaction terms to capture the combined effect of slope and distance [@lovelace2017propensity].

The potential demand calculations show that the current and potential number of cyclists both follow a bell-shaped distribution, with the number of trips peaking around the 3-4km commuting distance and then going back down for longer distances (see Figure \@ref(fig:potdemhistograms)).

@@

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="75%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

The above points are accounted for by using a weighted road network for routing. This has previously been done by multiplying all road segments without cycling infrastructure by an impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments proportional to the investment cost of bringing them to an acceptable level of stress for cycling  [@duthie2014optimization]. Our approach is similar to the latter, as we create a weighting profile that is adjusted to favor less stressful streets (based on information from Table  \@ref(table:osmroadtypes)), and roads with existing cycling infrastructure. We believe this to be more practical  than the approach adopted by @mauttone2017bicycle, as it goes beyond simply favoring roads with existing cycling infrastructure to creating a hierarchy of road preference based on perceived stress levels. The approach is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.

>>>>

<!-- - Intro: possible trim down substantially then add some literature in (a) Calculating Potential Demand [DONE]/ (b) Routing [DONE]/ (c) Road Segment Prioritization -->
<!-- - Community Detection: Where in the document should this section be? It is now before Road Segment Prioritization, but I can merge them -->
<!-- Can go into road segment section. -->
<!-- - Should we use other cities for the sake of comparison (compare communities, road types etc) -->
- How to add tables properly (they are now added using latex code which only shows up when outputting to pdf)
- Footnotes are sometimes split between pages. How do I keep them on the same page?
<!-- - What is the best way to write pseudocode (Steps for Algorithm 1 and 2) -->

@@

*Optimization* techniques have been used to propose improvements to cycling networks. @mesbah2012bilevel propose a bi-level formulation to optimize allocation of cycling lanes to the network without exceeding a set budget. They account for the effect of cycling lanes on car traffic, and attempt to maximize utilization of said lanes with minimal impact on car travel times. Car usage is not considered by @mauttone2017bicycle, who develop an optimization framework that aims to minimize the total user cost of cycling on the network. The aggregate flow on links^[*link* refers to a road segment throughout this research] is obtained by using shortest paths to route existing cycling demand onto the road network, and the solution is a proposed set of links where cycling infrastructure should be added in order to minimize the overall travel cost of cyclists across the network. The cost of traversing a link is given as a function of its length and whether or not it has cycling infrastructure. The problem has also been solved by attempting to find the minimum cost of improving roadway links to meet a desired level of service (LOS) [@duthie2014optimization]. In this formulation, all origin-destination (OD) pairs need to be connected by roads that meet the desired LOS. A limitation of these problem formulations do not explicitly solve for continuity, which is dealt with using either a constraint specifying that each link with a bike lane should be connected to at least one destination [@mesbah2012bilevel], a constraint on maximum deviation from shortest paths [@duthie2014optimization], or a discontinuity penalty to prioritize connected road segments [@mauttone2017bicycle].

@@

*Origin-Destination (OD)* data can be obtained from a range of sources, including GPS data, household travels surveys and Census data on work locations.
In areas where observed OD data is unavailable, modelling techniques such as spatial interaction models can be used to estimate travel volumes between zones [@ref?].
In this paper we use open access data from the UK census [@ONS2011flowdata], which contains aggregate statistics on number of commuters between administrative zones --- Middle layer Super Output Areas (MSOA) --- by mode of travel. MSOAs have an average population of just over 8000 [@ofn2018population].^[
See https://wicid.ukdataservice.ac.uk/ for open access to the OD data.
]
Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling) illustrates the proportion of trips cycled by distance and the geographic extent of the input OD dataset used in this paper.

@@

\noindent where **d** and **s** are the distance and slope respectively for the OD pair. The authors use square and square-root distance terms "to capture the non-linear impact of distance on the likelihood of cycling", and interaction terms to capture the combined effect of slope and distance [@lovelace2017propensity].
Alternative cycling uptake models could be used in different contexts or scenarios of change.

The potential demand calculations show that the current and potential number of cyclists both follow a bell-shaped distribution, with the number of trips peaking around the 3-4km commuting distance and then going back down for longer distances (see Figures \@ref(fig:potdemhistograms) and \@ref(fig:desirefacetcycling)).

<!-- note: could be a joint table/map showing sample - if reviewers request -->

@@

```{r desirefacetcycling, fig.cap="Current and Potential Cycling Demand", out.width="100%", fig.show='hold', echo=FALSE}
knitr::include_graphics("data/Manchester/Plots/desire_facet_cycling.png")
```

@@

The above points are accounted for by using a weighted road network for routing. This has previously been done by multiplying all road segments without cycling infrastructure by a fixed impedance factor [@mauttone2017bicycle], or by assigning different weights to the road segments proportional to the investment cost of bringing them to an acceptable level of stress for cycling  [@duthie2014optimization].

For the purposes of this paper, created a weighting profile that is adjusted to favor less stressful streets (based on information from Table  \@ref(table:osmroadtypes)), and roads with existing cycling infrastructure. We believe this to be more practical  than the simplistic approach adopted by @mauttone2017bicycle, as it goes beyond simply favoring roads with existing cycling infrastructure enabling a hierarchy of road preference based on perceived stress levels.
A wide range of weighting profiles could be used, to represent different types of cyclists and road environments, as described in @ref.
<!-- The approach is also in line with the creation of LTNs, as residential streets are those where motorized traffic is most likely to be banned in the creation of LTNs.  -->
